A molecular dynamics study of the solvation of carbon dioxide in the ionic liquids \ce{[emim][B(CN)_4]} and \ce{[emim][NTf_2]}

### 2018-09-07 ionic_liquids.tex
Revision of Activity Coefficients subsection

In what follows, the subscripts ``s'' and ``IL'' refer to the solute and the solvent (ionic liquid), respectively.
For a liquid phase containing $N_\text{\tiny IL}$ solvent molecules and $N_{\text{s}}$ solute molecules at fixed pressure (P) and temperature (T), the difference in free energy between the states associated to $\lambda$ = 1 and $\lambda$ = 0 is formally defined as
\begin{equation}
\Delta G_\text{sim} \equiv k_b T \ln \frac{\Xi (N_s,N_\text{\tiny IL},P,T,\lambda = 1)}{\Xi (N_s,N_\text{\tiny IL},P,T,\lambda = 0)}, 
\end{equation}
where $\Xi$ denotes the isobaric-isothermal partition function and $k_b$ is the Boltzmann constant.
Shirts \textit{et al}. \cite{Shirts_2003} rigorously derived the working expression that relates $\Delta G_\text{sim}$ to what is experimentally measured as the solvation free energy, namely
\begin{equation}
\label{eq:solvation free energy definition}
\Delta G_{\text{solv}} \equiv k_b T \ln \left( \frac{P_s}{c_s R T} \right),
\end{equation}
where $R = N_A k_b$ is the gas constant,
$N_A$ is the Avogadro number,
$P_s$ is the partial pressure of the solute in the gas phase, and
$c_s$ is the molar concentration of the solute in the liquid phase.
Note that these quantities correspond to a phase equilibrium condition.
For obtaining the mentioned relation, the authors derived an expression for the chemical potential of the solute in the liquid phase ($\mu^l_s$) as a function of $\Delta G_\text{sim}$, which can be written as
\begin{equation}
\mu^l_s = \mu^{ig}_s + \Delta G_\text{sim} - k_bT \ln \left( \frac{V^{\ast}}{V_1}\right) - k_bT \ln \left( \frac{P_s}{c_s R T} \right),
\end{equation}
where $\mu^{ig}_s$ is the chemical potential of the solute in an ideal gas at temperature $T$ and partial pressure of solute $P_s$, while $V^{\ast}$ and $V_1$ are the mean volumes of liquid solutions specified by ($N_s-1$,$N_\text{\tiny IL}$,P,T) and ($N_s$,$N_\text{\tiny IL}$,P,T), respectively.
Note that $c_s = \frac{N_s}{N_A V_1}$.
For small solutes, such as those considered in this work, the term containing the ratio $V^{\ast}/V_1$ is typically in the order of magnitude of the free-energy uncertainties \cite{Shirts_2003}.
Hence, it can be neglected and will be omitted hereafter.

@@

If both $P_s$ and $c_s^{\ast}$ are small, we can resort to Eqs.~\eqref{eq:solvation free energy definition} and \eqref{eq:free_solv} to predict Henry's law constant $\text{K}_s$, which is defined as \cite{Prausnitz}
\begin{equation}
\text{K}_s =\frac{P_s}{x_s}.
\end{equation}

Therefore, by simply expressing Eq.~\eqref{eq:solvation free energy definition} in terms of the molar fraction of the solute in the liquid phase ($x_s$) and assuming that the molar density of the solution does not perceivably differ from that of the solvent ($\tilde{\rho} \approx \tilde{\rho}_\text{\tiny IL}$), we obtain that
\begin{equation}
\label{eq:henry_eq}
K_s = \tilde{\rho}_\text{\tiny IL} R \, T \exp\left(\frac{\Delta G_\text{sim}}{k_b T}\right).
\end{equation}

@@

\subsection*{Infinite-dilution activity coefficient}

As will be shown below, the prediction of activity coefficients requires extra simulations involving the pure solute.
In what follows, we use the superscript ``$0$'' to denote the pure solute properties.

Using standard thermodynamic relations \cite{Tester}, we write the activity coefficient $\gamma_s$ in terms of residual chemical potentials, as follows:
\begin{equation}
\label{eq:gamma}
\ln \gamma_s = \frac{1}{RT} (\mu_s^R - \mu_s^{R,0}),
\end{equation}
where $\mu_s^R$ and $\mu_s^{R,0}$ are the residual chemical potentials of the solute in the solution and in the pure solute, respectively.
It is worth noting that we have used the definition $\mu_s^{R}$ (P,T) = $\mu_s$ (P,T) - $\mu_s^{ig}$ (P,T), where the superscript $ig$ refers to a solution of ideal gases.
Also, recall that $\mu_s^{ig} (P,T)$ = $\mu_s^{ig,0} (P,T)$ + $RT \ln x_s$, so we have that
\begin{equation}
\mu_s^R = \mu_s^l(N_s^l,N_{LI},P,T) - \mu_s^{ig,0}(P,T) - k_b T \ln x_s.
\end{equation}
Replacing $\mu_s^{l}$ and $\mu_s^{ig,0}$ by the corresponding expressions derived in Ref.~\cite{Shirts_2003} yields
\begin{equation}
\label{eq:mu_res}
\mu_s^R = \Delta G_\text{sim} - k_b T \ln \left( \frac{V^{\ast} P}{N_s^l k_b T} \right) -  k_b T \ln x_s.
\end{equation}
Of course, an analogous equation holds for $\mu_s^{R,0}$:
\begin{equation}
\label{eq:mu_res_0}
\mu_s^{R,0} = \Delta G_\text{sim}^{0} - k_b T \ln \left( \frac{V^{\ast,0} P}{N_s^{l,0} k_b T} \right).
\end{equation}

If one makes $N_s^{l}$ = 1, and substitute Eqs.~\eqref{eq:mu_res} and~\eqref{eq:mu_res_0} into Eq.~\eqref{eq:gamma}, the infinite-dilution activity coefficient, $\gamma^{\infty}_s$, can be predicted according to the following equation:
\begin{equation}
\label{eq:gamma_inf}
\ln \gamma^{\infty}_s = \frac{1}{k_b T} \left( \Delta G_\text{sim} - \Delta G_\text{sim}^{0} \right) + \frac{c_\text{\tiny IL}^{\, \ast}}{c_{\, s}^{\, \ast,0}},
\end{equation}
where $c_\text{\tiny IL}^{\, \ast}$ y $c_{\, s}^{\, \ast,0}$ are the numerical densities of the solvent and the pure solute, respectively.

>>>>

In what follows, the subscripts ``s'' and ``IL'' refer to the solute and the solvent (ionic liquid), respectively.
For a liquid phase containing $N_\text{\tiny IL}$ solvent molecules and $N_{\text{s}}$ solute molecules at fixed pressure (P) and temperature (T), the difference in free energy between the states associated to $\lambda$ = 1 and $\lambda$ = 0 is formally defined as
\begin{equation}
\Delta G_\text{sim} \equiv k_b T \ln \frac{\Xi (N_s,N_\text{\tiny IL},P,T,\lambda = 1)}{\Xi (N_s,N_\text{\tiny IL},P,T,\lambda = 0)}, 
\end{equation}
where $\Xi$ denotes the isobaric-isothermal partition function and $k_b$ is the Boltzmann constant.
Shirts \textit{et al}. \cite{Shirts_2003} rigorously derived the working expression that relates $\Delta G_\text{sim}$ to what is experimentally measured as the solvation free energy, namely
\begin{equation}
\label{eq:solvation free energy definition}
\Delta G_{\text{solv}} \equiv k_b T \ln \left( \frac{P_s}{c_s R T} \right),
\end{equation}
where $R = N_A k_b$ is the gas constant,
$N_A$ is the Avogadro number,
$P_s$ is the partial pressure of the solute in the gas phase, and
$c_s$ is the molar concentration of the solute in the liquid phase.
Note that these quantities correspond to a phase equilibrium condition.
For obtaining the mentioned relation, the authors derived an expression for the chemical potential of the solute in the liquid phase ($\mu^l_s$) as a function of $\Delta G_\text{sim}$, which can be written as
\begin{equation}
\label{eq:solute chemical potential}
\mu^l_s = \mu^{ig}_s + \Delta G_\text{sim} - k_bT \ln \left( \frac{V^{\ast}}{V_1}\right) - k_bT \ln \left( \frac{P_s}{c_s R T} \right),
\end{equation}
where $\mu^{ig}_s$ is the chemical potential of the solute in an ideal gas at temperature $T$ and solute partial pressure $P_s$, while $V^{\ast}$ and $V_1$ are the mean volumes of liquid solutions specified by ($N_s-1$,$N_\text{\tiny IL}$,P,T) and ($N_s$,$N_\text{\tiny IL}$,P,T), respectively.
Note that $c_s = \frac{N_s}{N_A V_1}$.
For small solutes, such as those considered in this work, the term containing the ratio $V^{\ast}/V_1$ is typically in the order of magnitude of the free-energy uncertainties \cite{Shirts_2003}.
Hence, it can be neglected and will be omitted hereafter.

@@

If both $P_s$ and $c_s$ are small, we can resort to Eqs.~\eqref{eq:solvation free energy definition} and \eqref{eq:free_solv} to predict Henry's law constant $\text{K}_s$, which is defined as \cite{Prausnitz}
\begin{equation}
\text{K}_s =\frac{P_s}{x_s}.
\end{equation}

Therefore, by simply expressing Eq.~\eqref{eq:solvation free energy definition} in terms of the molar fraction of the solute in the liquid phase ($x_s$) and assuming a dilute solution, whose molar density does not perceivably differ from that of pure solvent (that is, $\tilde{\rho} \approx \tilde{\rho}_\text{\tiny IL}$), we obtain that
\begin{equation}
\label{eq:henry_eq}
K_s = \tilde{\rho}_\text{\tiny IL} R \, T \exp\left(\frac{\Delta G_\text{sim}}{k_b T}\right).
\end{equation}

@@

\subsection*{Activity Coefficients}

As will be shown below, the prediction of activity coefficients requires additional simulations of the solute as a pure liquid phase, whose properties we denote here by a superscript ``$0$''.
We start by using standard thermodynamic relations \cite{Tester} to write the solute activity coefficient $\gamma_s$ in terms of chemical potentials as
\begin{equation}
\label{eq:gamma}
\ln \gamma_s = \frac{\mu_s^l - \mu_s^{l,0}}{k_B T} - \ln x_s,
\end{equation}

Next, we replace $\mu_s^l$ and $\mu_s^{l,0}$ by their corresponding instances of Eq.~\eqref{eq:solute chemical potential}. In this step, we note that $c_s = \tilde{\rho} x_s$ and $c_s^0 = \tilde{\rho}_s^0$, where $\tilde{\rho}$ and $\tilde{\rho}_s^0$ are the molar densities of the solution and of the pure liquid solute, respectively. Thus, the expression we obtain for computing an activity coefficient from two independent simulations is
\begin{equation}
\ln \gamma_s = \frac{\Delta G_\text{sim} - \Delta G_\text{sim}^{0}}{k_b T} + \ln \frac{\tilde{\rho}}{\tilde{\rho}_s^0}.
\end{equation}

In the special case of a solute at infinite dilution in an ionic liquid solvent, its activity coefficient $\gamma^\infty_s$ can be computed by replacing $\tilde{\rho}$ in the equation above by the molar density of the pure solvent, $\tilde{\rho}_\textit{\tiny IL}$.
